---
title: "Overlaps models with GBIF data for birds and mammals"
author: "Natalie Cooper"
date: "3/14/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This uses the overlaps data, plus region info and taxonomy from GBIF.

```{r, echo = FALSE, message = FALSE}
# Load libraries
library(tidyverse)
library(viridis)
library(knitr)
library(here)
library(patchwork)
library(ggfortify)
library(broom)

# Read in the overlaps data
ds <- read_csv(here("data/gbif-overlaps.csv"))

### Remember to delete duplicates if using region/continent not country ###
continents <- 
  ds %>%
  dplyr::select(-country, -region, -countrycode) %>%
  # Remove Peromyscus as it's a lot (1000+) higher than everything else.
  filter(NumberOverlaps < 1000) %>% 
  filter(clade != "Amphibians") %>%
  distinct()

# Helper functions for plotting
remove_y <- 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

First a quick summary of how the data are spread across clades and continents...
```{r, echo = FALSE}
sum1 <- 
ds %>%
  dplyr::select(-country, -region, -countrycode) %>%
  distinct() %>%
  group_by(clade, continent) %>%
  filter(clade != "Amphibians") %>%
  summarise(number = n())

kable(sum1)
```

### Overlaps by clade
Red dots are the mean with standard error bars, purple are medians, grey dots are the raw data.

```{r, echo = FALSE, warning = FALSE}
sum2 <- 
  continents %>%
  group_by(clade) %>%
  summarise(mn = mean(PercentOverlaps),
            med = median(PercentOverlaps),
            se = sqrt((var(PercentOverlaps)/length(PercentOverlaps))),
            totalo = sum(NumberOverlaps)/sum(NumberSpecimens)) 

ggplot(continents, aes(x = clade, y = PercentOverlaps)) +
  geom_point(alpha = 0.5, col = "grey", size = 0.5) +
  geom_point(data = sum2, aes(y = mn), col = "red", size = 3) +
  geom_point(data = sum2, aes(y = med), col = "purple", size = 3) +
  geom_errorbar(data = sum2, aes(y = mn, ymin = mn - se, ymax = mn + se), width = 0.1) +
  theme_bw(base_size = 14) +
  xlab("")
```

```{r}
sum2
```

### Overlaps by continent
Red dots are the mean with standard error bars, median is purple, other coloured dots are the raw data.
```{r, echo = FALSE, warning = FALSE}
sum3 <- 
  continents %>%
  group_by(continent) %>%
  summarise(mn = mean(PercentOverlaps),
            se = sqrt(var(PercentOverlaps)/length(PercentOverlaps)),
            med = median(PercentOverlaps))

ggplot(continents, aes(x = continent, y = PercentOverlaps, col = clade)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_point(data = sum3, aes(y = mn, col = NULL), col = "red", size = 3) +
  geom_point(data = sum3, aes(y = med, col = NULL), col = "purple", size = 3) +
  geom_errorbar(data = sum3, aes(y = mn, ymin = mn - se, ymax = mn + se, col = NULL), width = 0.1) +
  theme_bw(base_size = 14) +
  xlab("") +
  theme(legend.title = element_blank())
```

```{r}
sum3
```

### Percent overlaps

```{r, echo = FALSE, warning = FALSE}
p1 <-
ggplot(continents, aes(x = PercentOverlaps, fill = clade, col = clade)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "left")

p2 <-
  ggplot(continents, aes(x = PercentOverlaps, fill = continent, col = continent)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  remove_y +
  theme(legend.title = element_blank(),
        legend.position = "right")

p1 + p2
```

### Number of overlaps 

```{r, echo = FALSE, warning = FALSE}
p1 <-
ggplot(continents, aes(x = NumberOverlaps, fill = clade, col = clade)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "left")

p2 <-
  ggplot(continents, aes(x = NumberOverlaps, fill = continent, col = continent)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  remove_y +
  theme(legend.title = element_blank(),
        legend.position = "right")

p1 + p2
```

### Number of specimens 

```{r, echo = FALSE, warning = FALSE}
p1 <-
ggplot(continents, aes(x = NumberSpecimens, fill = clade, col = clade)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "left")

p2 <-
  ggplot(continents, aes(x = NumberSpecimens, fill = continent, col = continent)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  remove_y +
  theme(legend.title = element_blank(),
        legend.position = "right")

p1 + p2
```

### Is there are strong relationship between % overlap and number of specimens?

```{r}
ggplot(continents, aes(x = NumberSpecimens, y = PercentOverlaps, col = clade)) +
  geom_point(alpha = 0.5) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank(),
        legend.position = "right")
```


### Models

####1. By clade. GLM with quasibinomial errors. 

```{r, echo = FALSE, warning = FALSE}
ggplot(continents, aes(x = clade, y = NumberOverlaps)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw(base_size = 14) +
  xlab("") +
  theme(legend.title = element_blank())

```

```{r}
# Fitting the model 
model1 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ clade,
              data = continents, family = "binomial")
# Check for overdispersion (should be < 2)
sum_model1 <- summary(model1)
sum_model1$deviance / sum_model1$df.resid 

# Use quasibinomial model instead
model1 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ clade,
              data = continents, family = "quasibinomial")

# Look at model diagnostics
autoplot(model1)

# Look at outputs
tidy(anova(model1, test = "F"))
summary(model1)
```

####2. By continent. GLM with quasibinomial errors. 
Significant.

```{r, echo = FALSE, warning = FALSE}
ggplot(continents, aes(x = continent, y = NumberOverlaps, col = clade)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw(base_size = 14) +
  xlab("") +
  theme(legend.title = element_blank())
```

```{r}
# Fitting the model 
model2 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ continent,
              data = continents, family = "binomial")
# Check for overdispersion (should be < 2)
sum_model2 <- summary(model2)
sum_model2$deviance / sum_model2$df.resid 

# Use quasibinomial model instead
model2 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ continent,
              data = continents, family = "quasibinomial")

# Look at model diagnostics
autoplot(model2)

# Look at outputs
tidy(anova(model2, test = "F"))
summary(model2)
```

### What about body size?

```{r}
# Read in data and select required columns then rename for joins
mam <- read_delim(here("data/MamFuncDat.txt"), delim = "\t")

mam <-
  mam %>%
  rename(Binomial = Scientific) %>%
  dplyr::select(Binomial, 'BodyMass-Value')

bird <- read_delim(here("data/BirdFuncDat.txt"), delim = "\t")

bird <-
  bird %>%
  rename(Binomial = Scientific) %>%
  dplyr::select(Binomial, 'BodyMass-Value')

bm <- rbind(bird, mam)

continents2 <- left_join(continents, bm) %>%
  rename(bodymass = 'BodyMass-Value')
```

```{r, echo = FALSE, warning = FALSE}
ggplot(continents2, aes(x = log(bodymass), y = NumberOverlaps, col = clade)) +
  geom_point(alpha = 0.5, size = 0.5) +
  theme_bw(base_size = 14) +
  theme(legend.title = element_blank()) 
```

```{r}
# Fitting the model 
model3 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ log(bodymass)*clade,
              data = continents2, family = "binomial")
# Check for overdispersion (should be < 2)
sum_model3 <- summary(model3)
sum_model3$deviance / sum_model3$df.resid 

# Use quasibinomial model instead
model3 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ log(bodymass)*clade,
              data = continents2, family = "quasibinomial")

# Look at model diagnostics
autoplot(model3)

# Look at outputs
tidy(anova(model3, test = "F"))
summary(model3)
```

# IUCN Red List status

```{r, echo = FALSE, warning = FALSE}
# Read in iucn data and merge
iucn <- read_csv(here("data/IUCN-data.csv"))
all <- 
  continents %>%
  left_join(iucn, by = c("Binomial" = "scientific_name")) %>%
  mutate(category = factor(category, levels = c("LC", "NT", "VU", "EN", "CR", "EX", "DD"))) %>%
  filter(!is.na(category))
```

### Overlaps by IUCN status
Red dots are the mean with standard error bars, median is purple, other coloured dots are the raw data.
```{r, echo = FALSE, warning = FALSE}
sum4 <- 
  all %>%
  group_by(category) %>%
  summarise(mn = mean(PercentOverlaps),
            se = sqrt(var(PercentOverlaps)/length(PercentOverlaps)),
            med = median(PercentOverlaps),
            n())

ggplot(all, aes(x = category, y = PercentOverlaps, col = clade)) +
  geom_point(alpha = 0.5, size = 0.5) +
  geom_point(data = sum4, aes(y = mn, col = NULL), col = "red", size = 3) +
  geom_point(data = sum4, aes(y = med, col = NULL), col = "purple", size = 3) +
  geom_errorbar(data = sum4, aes(y = mn, ymin = mn - se, ymax = mn + se, col = NULL), width = 0.1) +
  theme_bw(base_size = 14) +
  xlab("") +
  theme(legend.title = element_blank())
```

```{r}
sum4
```
```{r}
  ggplot(all, aes(x = PercentOverlaps, fill = category, col = category)) +
  geom_histogram(alpha = 0.5, bins = 10) +
  theme_bw(base_size = 14) +
  remove_y +
  theme(legend.title = element_blank(),
        legend.position = "right")
```


# Models 

```{r}
# Fitting the model 
model4 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ category*clade,
              data = all, family = "binomial")
# Check for overdispersion (should be < 2)
sum_model4 <- summary(model4)
sum_model4$deviance / sum_model4$df.resid 

# Use quasibinomial model instead
model4 <- glm(cbind(NumberOverlaps, (NumberSpecimens - NumberOverlaps)) ~ category,
              data = all, family = "quasibinomial")

# Look at model diagnostics
autoplot(model4)

# Look at outputs
tidy(anova(model4, test = "F"))
summary(model4)
```




```{r}
anova(model4, test = 'Chisq')
```



